---
title: "RNA-Seq Exploratory Data Analysis"
author: "Your Name"
format:
  html:
    self-contained: true
  pdf:
    toc: true
    number-sections: true
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
---

# EDA On Your Own

Now that you have taken a guided tour of EDA for RNA-seq datasets, I challenge you to do the entire thing again on your own.

I want you to explore [stomach]{.underline} cancer cell lines in CCLE. How many biologically distinct groups can you reasonably define? What do the gene expression profiles of each group tell us about the biology of each group?

# Load Libraries

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(scales)
library(ComplexHeatmap)
library(EnhancedVolcano)
library(DESeq2)
```

# Prepare Data

```{r}
ccle_counts <- read.csv("/cloud/project/outs/1_ccle_counts.csv", row.names = 1)
ccle_meta <- read.csv("/cloud/project/outs/1_ccle_meta.csv")
```

```{r}
unique(ccle_meta$Site_Primary)
```

```{r}
index_a <- ccle_meta$Site_Primary %in% c("stomach")
index_b = ccle_meta$Pathology != ''
ccle_counts_subset <- ccle_counts[,index_a & index_b]
ccle_meta_subset <- ccle_meta[index_a & index_b,]
```

```{r}
table(ccle_meta_subset$tcga_code)
```

# Principal Component Analysis

```{r}
dds <- DESeqDataSetFromMatrix(countData = data.frame(ccle_counts_subset),
                              colData = ccle_meta_subset,
                              design = ~ Pathology)
dds <- dds[rowSums(counts(dds)) >= 100,]
vsd <- vst(dds, blind = FALSE)
```

```{r}
variances <- rowVars(assay(vsd))
top_var_genes <- order(variances, decreasing = TRUE)[1:500]
high_var_mat <- assay(vsd)[top_var_genes,]
pca_res <- prcomp(t(high_var_mat), center = TRUE, scale = FALSE)

explained_variance <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

# PC1 vs PC2
df_pc12 <- data.frame(PC1 = pca_res$x[, 1], PC2 = pca_res$x[, 2], ccle_meta_subset)

ggplot(df_pc12, aes(x = PC1, y = PC2, color = Pathology)) +
  geom_point(size = 3, alpha = 0.5) +
  theme_classic() +
  labs(x = paste0("PC1: ", round(explained_variance[1], 1), "% variance"), 
       y = paste0("PC2: ", round(explained_variance[2], 1), "% variance"))

# PC3 vs PC4
df_pc34 <- data.frame(PC3 = pca_res$x[, 3], PC4 = pca_res$x[, 4], ccle_meta_subset)

ggplot(df_pc34, aes(x = PC3, y = PC4, color = Pathology)) +
  geom_point(size = 3, alpha = 0.5) +
  theme_classic() +
  labs(x = paste0("PC3 (", round(explained_variance[3], 2), "%)"), 
       y = paste0("PC4 (", round(explained_variance[4], 2), "%)"))

# Elbow plot
df_variance <- data.frame(PC = seq_len(25), Variance = explained_variance[1:25])

ggplot(df_variance, aes(x = PC, y = Variance)) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Elbow Plot (Top 25 PCs)", x = "Principal Component", y = "Explained Variance (%)")
```

```{r}
variances <- rowVars(assay(vsd))
top_var_genes <- order(variances, decreasing = TRUE)[1:500]
high_var_mat <- assay(vsd)[top_var_genes,]
pca_res <- prcomp(t(high_var_mat), center = TRUE, scale = FALSE)
df_pca <- data.frame(PC1 = pca_res$x[, 1], PC2 = pca_res$x[, 2], ccle_meta_subset)

pc1_loadings <- pca_res$rotation[, 1]
pc1_sorted_loadings <- sort(pc1_loadings, decreasing = TRUE)
pc1_top_genes <- names(head(pc1_sorted_loadings, n = 6))
pc1_top_genes_vsd <- data.frame(t(assay(vsd)[pc1_top_genes, ]))

df_genes_long <- data.frame(
  pc1_top_genes_vsd,  
  df_pca[, c("PC1", "PC2", "Pathology")]
)

df_long <- pivot_longer(
  df_genes_long,
  cols = all_of(pc1_top_genes),
  names_to = "Gene",
  values_to = "Expression"
)

ggplot(df_long, aes(x = PC1, y = PC2, color = Expression)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_classic() +
  labs(x = paste0("PC1: ", round(explained_variance[1], 1), "% variance"), 
       y = paste0("PC2: ", round(explained_variance[2], 1), "% variance"),
       color = "Expression (VST)") +
  facet_wrap(~ Gene)
```

```{r}
pc2_loadings <- pca_res$rotation[, 2]
pc2_sorted_loadings <- sort(pc2_loadings, decreasing = TRUE)
pc2_top_genes <- names(head(pc2_sorted_loadings, n = 6))
pc2_top_genes_vsd <- data.frame(t(assay(vsd)[pc2_top_genes, ]))

df_genes_long <- data.frame(
  pc2_top_genes_vsd,  
  df_pca[, c("PC1", "PC2", "Pathology")]
)

df_long <- pivot_longer(
  df_genes_long,
  cols = all_of(pc2_top_genes),
  names_to = "Gene",
  values_to = "Expression"
)

ggplot(df_long, aes(x = PC1, y = PC2, color = Expression)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_classic() +
  labs(x = paste0("PC1: ", round(explained_variance[1], 1), "% variance"), 
       y = paste0("PC2: ", round(explained_variance[2], 1), "% variance"),
       color = "Expression (VST)") +
  facet_wrap( ~ Gene)
```

# Hierarchical Clustering

```{r}
high_var_genes <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 500)
high_var_mat <- assay(vsd)[high_var_genes,]

hc <- hclust(dist(t(high_var_mat)), method = "complete")
clusters <- cutree(hc, k = 2)

Heatmap(high_var_mat,
        clustering_distance_columns = "euclidean",
        clustering_method_columns = "complete",
        clustering_distance_rows = "euclidean",
        clustering_method_rows = "complete",
        show_row_names = FALSE,
        show_column_names = FALSE,
        name = "VST_Counts",
        column_split = clusters)

pca_res <- prcomp(t(high_var_mat), center = TRUE, scale = FALSE)
explained_variance <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100
df_pca <- data.frame(PC1 = pca_res$x[, 1], PC2 = pca_res$x[, 2], ccle_meta_subset)
df_pca$Cluster <- factor(clusters)

ggplot(df_pca, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 3, alpha = 0.5) +
  scale_color_manual(values = hue_pal()(length(unique(clusters)))) +
  theme_classic() +
  labs(x = paste0("PC1: ", round(explained_variance[1], 1), "% variance"), 
       y = paste0("PC2: ", round(explained_variance[2], 1), "% variance"),
       color = "Cluster")
```

# Differential Expression Analysis

```{r}
high_var_genes <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 500)
high_var_mat <- assay(vsd)[high_var_genes,]

hc <- hclust(dist(t(high_var_mat)), method = "complete")
clusters <- cutree(hc, k = 2)
ccle_meta_subset$Cluster <- factor(clusters)

dds <- DESeqDataSetFromMatrix(countData = data.frame(ccle_counts_subset),
                              colData = ccle_meta_subset,
                              design = ~ Cluster)
dds$Cluster <- relevel(dds$Cluster, ref = "2")
dds <- dds[rowSums(counts(dds)) >= 100,]
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
print(head(res, 20))
```

```{r}
summary(res)
```

```{r}
res_reordered <- res %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "gene_symbol") %>%
  filter(padj < 0.001) %>%
  arrange(desc(abs(log2FoldChange)))

print(head(res_reordered, 20))
```

```{r}
degs <- c("MUC17", "LYZ", "REG4", "MAGEA10", "ADAMTS12", "NPFFR2")
degs_vsd <- data.frame(t(assay(vsd)[degs, ]))

variances <- rowVars(assay(vsd))
top_var_genes <- order(variances, decreasing = TRUE)[1:500]
high_var_mat <- assay(vsd)[top_var_genes,]
pca_res <- prcomp(t(high_var_mat), center = TRUE, scale = FALSE)
df_pca <- data.frame(PC1 = pca_res$x[, 1], PC2 = pca_res$x[, 2], ccle_meta_subset)

df_genes_long <- data.frame(
  degs_vsd,
  df_pca[, c("PC1", "PC2", "tcga_code")]
)

df_long <- pivot_longer(
  df_genes_long,
  cols = all_of(degs),
  names_to = "Gene",
  values_to = "Expression"
)

ggplot(df_long, aes(x = PC1, y = PC2, color = Expression)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_classic() +
  labs(x = paste0("PC1: ", round(explained_variance[1], 1), "% variance"), 
       y = paste0("PC2: ", round(explained_variance[2], 1), "% variance"),
       color = "Expression (VST)") +
  facet_wrap(~ Gene)
```

```{r}
top_degs_up <- res_reordered %>%
  arrange(desc(log2FoldChange)) %>%
  head(10)

top_degs_down <- res_reordered %>%
  arrange(log2FoldChange) %>%
  head(10)

genes_to_label <- c(top_degs_up$gene, top_degs_down$gene)

EnhancedVolcano(res_reordered,
                x = "log2FoldChange",
                y = "padj",
                lab = res_reordered$gene,
                selectLab = genes_to_label,
                pCutoff = 0.00001,
                FCcutoff = 2,
                labSize = 3,
                title = "Cluster 1 vs Cluster 2 DEGs",
                subtitle = "",
                caption = "",
                pointSize = 1,
                colAlpha = 0.3,
                drawConnectors = TRUE,
                widthConnectors = 0.5,
                max.overlaps = Inf
                ) +
  theme_classic() +
  theme(legend.position = "none")
```

# Interpretation

How many biologically distinct groups of stomach cancer cell lines did you define? What do the gene expression profiles of each group tell us about the biology of each group?

```{r}
# Your written response here

# The gene expression profile of Cluster 1 indicates involvement of:
  # MAGEA10, MAGEA4 are known cancer/testis antigens.
  # Cell migration, extracelluar matrix remodeling, epithelial-mesenchymal transition (EMT): WNT5A, BASP1, CRISPLD1, KANK4, ADAMTS12, ROBO2
  # Signaling pathways related to cell growth and survival: NPFFR2

# The gene expression profile of Cluster 2 indicates involvement of:
  # Mucosal immunity and inflammation: LYZ, REG4, and MUC17 are involved in mucosal defense and protection, which are often activated in gastrointestinal cancer.
  # Cell adhesion and metastasis: CEACAM5 and AGR3 are involved in adhesion and cell-cell interactions, likely influencing cancer metastasis.
  # LGALS4 is involved in modulating the tumor microenvironment, influencing cell adhesion and immune response, particularly by regulating T-cell activation
```

# sessionInfo()

```{r}
sessionInfo()
```
